Does an earthquake “know” how big it will be? A neural-net aided study

### 2024-06-02 article.tex
Captions and some figure adaptations.

Our analysis reveals a persistent increase in information gain across the test set for all investigated regions (Fig. \ref{fig:info_gain_over_time}b-d). This includes regions previously studied where no information gain was detected using different methods  \cite{ogata_exploring_2018}. Similar domain representations for additional regions can be found in the supplementary material  \neri{TODO: add these colorful plots for NZ and JAPAN to the SM}. Furthermore, the smoothed instantaneous information gain per event, presented in Fig. \ref{fig:info_gain_over_time}e-g, consistently exceeds the threshold for zero information gain marked by the red dashed line.

@@

For our analysis, we set $\Tilde{m} = m_c(t)$, determined dynamically using the maximum curvature method \cite{wiemer_minimum_2000} within a window of 150 past and 150 current and future events. The resulting temporal incompleteness curves are presented in the Extended Material Fig. \ref{fig:temp_incompleteness}. After applying the renormalization of Eq. \ref{eq:conditioned_likelihood} the cumulative information gain is recalculated. The result is presented as the grey dashed curved in Fig.  \ref{fig:info_gain_over_time}a, as well as in Fig. \ref{fig:info_gain_over_time}b-d for all three tested regions. Evidently, all conditioned cumulative information gain curves show a steadily increasing trend, supporting the hypothesis that there is information content about the upcoming magnitude in the statistical properties of the catalog, excluding the temporal incompleteness artifact. A bar plot comparison of the conditioned mean LL scores are presented in Extended Material Fig. \ref{fig:mean_ll_mc_conditioned}. Extended Material Table \ref{tab:mean_ll_all_benchmarks_conditioned} displays the conditioned scores for all benchmarks tested in this study. \neri{Perhaps the barplot is unecessary?}

@@

@@

\section*{Data and code availability}
The datasets analysed during the current study are available in the relevant references mentioned throught the article.
The code for analysing the data is available in \textcolor{red}{\textbf{**link TBD**}}
% This is a required section. Guidelines for data availability: \url{https://www.nature.com/documents/nr-data-availability-statements-data-citations.pdf}.

\newpage
\renewcommand\refname{Methods References}
\begin{thebibliography}{10}
\bibitem{zlydenko_neural_2023}
Zlydenko, O. et al. A neural encoder for earthquake rate forecasting. Sci Rep 13, 12350 (2023).
\bibitem{taroni_highdefinition_2021}
Taroni, M., Zhuang, J. \& Marzocchi, W. High‐Definition Mapping of the Gutenberg–Richter b‐Value and Its Relevance: A Case Study in Italy. Seismological Research Letters 92, 3778–3784 (2021).

@@

\newpage
\begin{figure}[h!]
	\centering
        \includegraphics[width=1\textwidth]{figures/info_gain_scatter_NZ_and_Japan.pdf}
	\caption{
            \textbf{Information gain of MAGNET over the benchmark, New Zealand and Japan data sets.} Scattered dots indicate magnitude at event index, colored by the information gain per event. \textbf{a} Entire test set of New Zealand and a selected time span (\textbf{b}) focusing on the 2016 Te Araroa earthquake and the 2016 Kaikōura earthquake. \textbf{c} The entire Japan data set and a focused time span \textbf{d} of the 2011 Tōhoku earthquake and some of its aftershocks.
         }
\label{fig:nz_japan_info_gain}
\end{figure}

\newpage
\begin{figure}[h!]
    \centering
        \includegraphics[width=1\textwidth]{figures/temporal_incompleteness.pdf}
    \caption{
    \textbf{Temporal Incompleteness for test sets of all regions examined.} Calculated temporal incompleteness for test sets is presented in red, superimposed over the events above (black) and below (grey) the completeness magnitude of the train set, $m_c^{(train)}$. Events are plotted by the index of their appearance in the test set. Events of low magnitude, $m_c^{(train)}$ (grey), that are not included in the evaluation, are plotted at their interpolated location along the horizontal axis. Data for Japan (c) is displayed only near the 2011 Tōhoku earthquake to reduce data density and enhance clarity of the figure.
    }
    \label{fig:temp_incompleteness}
\end{figure}

\newpage
\begin{figure}[h!]
    \centering
    \includegraphics[width=1\textwidth]{figures/raw_results_em.pdf}
    \caption{
        write the caption
    }
    \label{fig:model_output_em}
\end{figure}

\newpage
\begin{figure}[h!]
    \centering
    \includegraphics[width=1\textwidth]{figures/likelihood_scatter.pdf}
    \caption{
        write the caption
    }
    \label{fig:labels_likelihood}
\end{figure}

\newpage
\begin{figure}[h!]
    \centering
    \includegraphics[width=1\textwidth]{figures/combined_barplots_conditioned.pdf}
    \caption{
        \textbf{mean ll mc conditioned}
    }
    \label{fig:mean_ll_mc_conditioned}
\end{figure}

\newpage
\begin{figure}[h!]
    \centering
    \includegraphics[width=1\textwidth]{figures/completeness_cond.pdf}
    \caption{
        \textbf{mean ll conditioned on spatial mc}
    }
    \label{fig:mean_ll_conditioned_spatial_mc}
\end{figure}

>>>>

Our analysis reveals a persistent increase in information gain across the test set for all investigated regions (Fig. \ref{fig:info_gain_over_time}b-d). This includes regions previously studied where no information gain was detected using different methods  \cite{ogata_exploring_2018}. Similar domain representations for additional regions can be found in the supplementary material. Furthermore, the smoothed instantaneous information gain per event, presented in Fig. \ref{fig:info_gain_over_time}e-g, consistently exceeds the threshold for zero information gain marked by the red dashed line.

@@

For our analysis, we set $\Tilde{m} = m_c(t)$, determined dynamically using the maximum curvature method \cite{wiemer_minimum_2000} within a window of 150 past and 150 current and future events. The resulting temporal incompleteness curves are presented in the Extended Material Fig. \ref{fig:temp_incompleteness}. After applying the renormalization of Eq. \ref{eq:conditioned_likelihood} the cumulative information gain is recalculated. The result is presented as the grey dashed curved in Fig.  \ref{fig:info_gain_over_time}a, as well as in Fig. \ref{fig:info_gain_over_time}b-d for all three tested regions. Evidently, all conditioned cumulative information gain curves show a steadily increasing trend, supporting the hypothesis that there is information content about the upcoming magnitude in the statistical properties of the catalog, excluding the temporal incompleteness artifact. Extended Material Table \ref{tab:mean_ll_all_benchmarks_conditioned} displays the conditioned scores for all benchmarks tested in this study.

@@

\subparagraph{Kernel density estimation (KDE) of last n events} Selecting the last $n$ events (similar to described in \textit{last $n$ events}) we use the computed KDE as a magnitude predictor for the current timestamp. The KDE is computed using using the SciPy tool\cite{scipy_2020} with 'scott' estimator bandwidth\cite{scott_2015}.

@@

\section*{Data and code availability}
The datasets analysed during the current study are available in the relevant references mentioned throught the article.
The code for analysing the data is available in \textcolor{red}{\textbf{**link TBD**}}

Country borders in this work are plotted using data from https://geojson-maps.kyd.au/ .
% This is a required section. Guidelines for data availability: \url{https://www.nature.com/documents/nr-data-availability-statements-data-citations.pdf}.

\newpage
\renewcommand\refname{Methods References}
\begin{thebibliography}{10}
\bibitem{zlydenko_neural_2023}
Zlydenko, O. et al. A neural encoder for earthquake rate forecasting. Sci Rep 13, 12350 (2023).
\bibitem{taroni_highdefinition_2021}
Taroni, M., Zhuang, J. \& Marzocchi, W. High‐Definition Mapping of the Gutenberg–Richter b‐Value and Its Relevance: A Case Study in Italy. Seismological Research Letters 92, 3778–3784 (2021).
\bibitem{scott_2015}
Scott, David W. Multivariate density estimation: theory, practice, and visualization. John Wiley \& Sons, 2015.  
\bibitem{scipy_2020}
Virtanen, P., Gommers, R., Oliphant, T.E. et al. SciPy 1.0: fundamental algorithms for scientific computing in Python. Nat Methods 17, 261–272 (2020).

@@

\newpage
\begin{figure}[h!]
	\centering
        \includegraphics[width=1\textwidth]{figures/info_gain_scatter_NZ_and_Japan.pdf}
	\caption{
            \textbf{Information gain of MAGNET over the benchmark, New Zealand and Japan data sets.} Scattered dots indicate magnitude at event index, colored by the information gain per event. \textbf{a} Entire test set of New Zealand and a selected time span (\textbf{b}) focusing on the 2016 Te Araroa earthquake and the 2016 Kaikōura earthquake. \textbf{c} The entire Japan data set and a focused time span \textbf{d} of the 2011 Tōhoku earthquake and some of its aftershocks. Low magnitude data points in a, c and d are subsampled to ease on image rendering.
         }
\label{fig:nz_japan_info_gain}
\end{figure}

\newpage
\begin{figure}[h!]
    \centering
        \includegraphics[width=1\textwidth]{figures/temporal_incompleteness.pdf}
    \caption{
    \textbf{Temporal Incompleteness for test sets of all regions examined.} Calculated temporal incompleteness for test sets is presented in red, superimposed over the events above (black) and below (grey) the completeness magnitude of the train set, $m_c^{(train)}$. Events are plotted by the index of their appearance in the test set. Events of low magnitude, $m_c^{(train)}$ (grey), that are not included in the evaluation, are plotted at their interpolated location along the horizontal axis. Data for Japan (c) is displayed only near the 2011 Tōhoku earthquake to reduce data density and enhance clarity of the figure. Low magnitude data points in b and c are subsampled to ease on image rendering.
    }
    \label{fig:temp_incompleteness}
\end{figure}

\newpage
\begin{figure}[h!]
    \centering
    \includegraphics[width=1\textwidth]{figures/raw_results_em.pdf}
    \caption{
        \textbf{MAGNET’s output for Japan and New Zealand data sets}. \textbf{a} (\textbf{c}), PDFs produced by MAGNET for 100 randomly sampled events from the test set of the New Zealand (Japan) data set, with the train set’s Gutenberg-Richter distribution superimposed (dashed black line). The true magnitude label, i.e. the true magntude of the event in query, is indicated by the color, interpreted by the colorbar in a. \textbf{b} (\textbf{d}), Marginal magnitude distribution, i.e. stationary p(m), for the New Zealand (Japan) data set. Histograms of the train and test sets are presented in orange and blue respectively.
    }
    \label{fig:model_output_em}
\end{figure}

\newpage
\begin{figure}[h!]
    \centering
    \includegraphics[width=1\textwidth]{figures/likelihood_scatter.pdf}
    \caption{
    \textbf{Log-likelihood of the true magnitude, $ell_i$}. The likelihood for the true magnitude label of the test set for each data set, $\ell_i=\log\left(p_{\textbf{x}_i, t_i}(m_i)\right)$, as defined in main text. $\ell_i$ values of MAGNET and the last 300 events benchmark are shown in red and orange respectively. The blue dashed line indicates the values that the GR benchmark would obtain.
    }
    \label{fig:labels_likelihood}
\end{figure}

% \newpage
% \begin{figure}[h!]
%     \centering
%     \includegraphics[width=1\textwidth]{figures/combined_barplots_conditioned.pdf}
%     \caption{
%         \textbf{mean ll mc conditioned}
%     }
%     \label{fig:mean_ll_mc_conditioned}
% \end{figure}

\newpage
\begin{figure}[h!]
    \centering
    \includegraphics[width=1\textwidth]{figures/completeness_cond.pdf}
    \caption{
        \textbf{Spatial effects on MAGNET's performance.} \textbf{a-d} Mean log likelihood scores, $\mathcal{L}$, of events contained in bins which hold $m_c(x,y) \leq m_c$, presented in \textbf{e-h}. The data set Japan polygon is a tight polygon surrounding the main cluster of low completeness bins in the Japan data set (detailed in main text). \textbf{i-l} display the full data set's $m_c(x,y)$ spatial distribution.
    }
    \label{fig:mean_ll_conditioned_spatial_mc}
\end{figure}
